From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction

### 2015-09-30 paper.tex
Rewritten abstract to improve coherence of the paper
Based on James' suggestions

Successfully sustaining face-to-face interaction between humans and robots
mandates fine perception of each other attentional focus. This article presents
a novel open-source technique to assess the visual focus of attention (VFoA)
based on accurate, real-time head pose estimation with a single monocular
camera.  Using manual video annotations as ground truth, we show that our system
reaches satisfactory levels of accuracy in a real-world face-to-face interaction
scenario.

We secondly propose to introduce the concept of ``with-me-ness'', borrowed from the
field of {\it Computer-Supported Collaborative Learning}, as an
\emph{in-the-moment} measurement of the quality of interaction. We argue that
this new metric, more specific than the broad concept of engagement, may be well
suited for operational, real-time assessment of an on-going interaction.

>>>>

Measuring ``how much the human is in the interaction'' -- the \emph{level of
engagement} -- is instrumental in building effective interactive robots.

Engagement, however, is a complex, multi-faceted cognitive mechanism that is
only indirectly observable. This article formalizes \emph{with-me-ness} as one
of such indirect measures. \emph{With-me-ness}, a concept borrowed from the
field of {\it Computer-Supported Collaborative Learning}, measures in a
well-defined way to what extend the human is \emph{with} the robot over the
course of an interactive task. As such, it is an meaningful precursor of
engagement.

We expose the full methodology, from real-time estimation of the human's focus
of attention (relying on a novel, open-source, vision-based head pose
estimator), to on-line computation of with-me-ness.

We report as well on the experimental validation of this approach, using a
naturalistic setup involving children during a complex robot-teaching task.
